A comparison of different methods for calculating tangent-stiffness matrices in a massively parallel computational peridynamics code.

### 2014-02-20 jacobian_calc_CMAME.tex
made additions render

%%
%% Copyright 2007, 2008, 2009 Elsevier Ltd
%%
%% This file is part of the 'Elsarticle Bundle'.
%% ---------------------------------------------
%%
%% It may be distributed under the conditions of the LaTeX Project Public
%% License, either version 1.2 of this license or (at your option) any
%% later version.  The latest version of this license is in
%%    http://www.latex-project.org/lppl.txt
%% and version 1.2 or later is part of all distributions of LaTeX
%% version 1999/12/01 or later.
%%
%% The list of all files belonging to the 'Elsarticle Bundle' is
%% given in the file `manifest.txt'.
%%

%% Template article for Elsevier's document class `elsarticle'
%% with numbered style bibliographic references
%% SP 2008/03/01
%%
%%
%%
%% $Id: elsarticle-template-num.tex 4 2009-10-24 08:22:58Z rishi $
%%
%%
\documentclass[preprint,12pt]{elsarticle}

@@

\note[Michael]{re: I.5, this is taken from the thesis.}
\add[Michael]{
\section{Working examples of CS, AD and FD for Newton's method}
\label{sec:WorkingExamples}
In order to illustrate the complex-step technique in a simpler form than it
appears in the \emph{Peridigm} code written for this study, an example
header-only library and two example problem definition programs that use it
have been created. All the materials described here are available in the
associated project repository.

@@

\begin{figure}[tbp]
  \centering
  \includegraphics[width=8cm]{./figs/truss.png}
  \caption{Beam and spring simulated in example program.}
  \label{fig:TrussSchematic}
\end{figure}

@@

Table \ref{tab:ConvergenceStudy} indicates that when using AD or the 'actual'
analytically derived Jacobian, a minimum number of iterations is achieved
corresponding to a maximum convergence rate. CS and FD are similarly affected
by a grossly large step-size, while subtractive cancellation invalidates the
results for FD for very small step-sizes. Inaccuracy for large step-sizes is is
predicted by the FD and CS formulas, which require a small
step-size to allow the neglect of higher order 12 terms. The results of this
test show that step-size selection is not unimportant when using FD, but when
properly configured any of the methods may return the same maximum convergence
rate for this problem.  Overall conclusions regarding the methods that the accuracy of the
Jacobian should not be confused with the accuracy of the equilibrium
solution, since each of the methods discussed here may be equally capable given
enough iterations. Numerical measurements of accuracy besides number of iterations taken
are not possible here because there is not basis for comparison, as the methods are solving
the problem separately and producing difference Newton iterates. Additionally, while the inaccuracy of the Jacobian may
strongly affect convergence rates, this relationship was not precisely quantified here.
}

@@

\note[Michael]{Re: I.5}
\add[Michael]{
\section{Convergence and displacement accuracy studies} 
\label{sec:PeridigmConvergenceStudy}
%
While the previous sections profiled the per-iteration performance of each of
the methods, it was necessary to conduct convergence rate studies in order to
get a complete picture. It is necessary for an analyist to know if the
accuracy differences measured in the per-iteration tests would amount to
differences in convergence rates among the methods if they were allowed to solve problems
normally. The question is posed, given that step-size $h$ is heuristically
selected by \emph{Perdigm} for the finite difference methods and presumable
makes a good choice, do the single-iteration Jacobian accuracy results predict
the ranking of the methods in terms of convergence rate for a given study? For
brevity, convergence rate was measured for each of the methods in solving just
the largest problem of the single-core test series. As was seen in the per-iteration results,
Jacobian accuracy did not vary greatly from test to test, such that any pair of tests
from the two series could have been considered reasonably representative. As in
the validation problem shown in ~\ref{subsec:Validation}, would each of the methods return an
equal convergence rate so long as $h$ were optimally selected?

@@

\subsection{Remarks}
Based on the experimental results, it appears that for users of \emph{Peridigm}
there is a case to be made for using AD for reasons of speed. However, certain
parallel tests done in this study showed a small performance gain over AD with
the FD method.  The results showed that CS produced more accurate
tangent-stiffness matrices than CD and FD under the parameters of the tests,
however it was determined that this difference in accuracy was not sufficient
to alter convergence rate for a test problem. In order for Jacobian accuracy to
have an impact on convergence rate, and therefor serve as a predictor of
convergence rate in per-iteration results, the amount of error detected between
a supposed perfectly accurate Jacobian and the test Jacobian needs to be
greater than allowed by the conditions seen in \cite{burkeLectures}, 
reproduced in \ref{sec:FurtherStudies}.

@@

However, at least for the models producing the results shown here, each of the
methods returned quadratic convergence, indicating each was sufficiently
accurate. 'Marching orders' for the analyst would be to select the swiftest
method, while it would seem accuracy above what is adequate does not enhance
the convergence rate or iteration speed until either classical Newton's convergence
conditions or the approximate case conditions are violated. 
}
The CS method was shown to be highly accurate in calculating tangent-stiffness
matrices by the standards of this study, and it is slower than other methods
for a naive implementation. In the context of future use of CS in
\emph{Peridigm}, CS holds the advantage of relying on byte-copyable data types,
while AD requires a complicated serialization and deserialization in order to
function as byte-copyable. This is fine for copying in between MPI processes,
but deserialization on an accelerator or GPU is very difficult, as this would
require compiling a version of the AD library being used that were compatible
with the accelerator or GPU being used. \note[Michael]{re:
I.3}\add[Michael]{For the reader's convenience, a collection of contiguously
stored data elements are said to be 'byte-copiable' when the the interval
in bytes of memory spanned by a data element in the collection is constant
throughout the collection. Suppose 'contiguously stored' means, stored in
memory with a minimum of unrelated infomation stored in between data
elements in the collection. Complicated objects, such as AD datatype
elements, do not have the same size in memory as other objects even of that
same class, for the reason that they themselves contain an uncertain
ammount of state data according to their design and work done in the
program. This means that a collection of such objects, even if they are
contiguously stored in memory, will not be alligned according to a constant
interval, such that elements within a collection cannot be located by
coordinates consisting of the location of the beginning of the first
element in the collection and an interval multiplied by a natural number.
This method of locating is integral to the serialized communication
implemented by MPI the libraries as well as CUDA, OpenCL, and others,
which allow only the transmission of collections of byte-copiable datatypes
(that is 'primitive datatypes'). A common method of bypassing this
limitation for communication purposes is to write the complicated data to a
'character buffer', and transmit the byte-copiable character buffer,
however this is useless unless the recipient device of the communication
can recover the original data and process it, which is impossible if the
recipient device only can work with a subset of the programming language
that the sending device must use, such as in a communication between a CPU
running C++ and a GPU running C (such as in CUDA programs), where the gap in
sophistication contains the features of the "complicated" datatype.} What this means is
that if \emph{Peridigm} or any simulation software were to be reformulated to
take advantage of accelerators or GPUs, the flexibilty of complex-step owed to
its simple implementation may make it more viable for that application given
its high accuracy. Additional performance gains may be achieved through the use
of the templated \emph{TPetra} package in \emph{Trilinos} which would allow the
use of complex data types natively, and alleviate the expense of the
dynamic-casts that were utilized in this study.  As a final comment, it should
be noted that from a code development perspective the CS method may be the
easiest of the methods to implement; It only requires one functional
evaluation, there is no heuristic choice of probe distance, and there is no
dependence on an external AD library.  From this perspective, the method may
still be an attractive method for TS matrix evaluation. \add[Michael]{However, at least for 
the models producing the results shown here, each of the
methods returned quadratic convergence, indicating each was sufficiently
accurate. 'Marching orders' for the analyst would be to select the swiftest
method, while it would seem accuracy above what is adequate does not enhance
the convergence rate or iteration speed until either classical Newton's convergence
conditions or the approximate case conditions are violated.}

@@

\item 
\label{differentiation}
%AD is equivalent to the chain rule, when the computer has definitions for functions and partials
%that can be adapted to match the composition function.
It follows from \ref{composition} that we can evaluate $\frac{d}{d x}f(x) \mid_{x_0}$ with the chain rule: \\ \\
$\frac{d}{d x}f(x) = \frac{d}{d x} \cdot g(h(k(x))) \mid_{x_0}$ \\
$\frac{d}{d x}f(x) =  \frac{d{g}}{d{h}} \cdot \frac{d{h}}{d{k}}
\cdot \frac{d{k}}{d{x}}\mid_{x_0}$ \\ \\
From the rest of \ref{given} it follows that we can approximate  $\frac{d}{d x}f(x) \mid_{x_0}$ by
specializing the computer's general forms of $u, v, w$ 
according to \ref{abcvalues}, with parameters $a, b$ chosen for each function and held as constant, and
evaluating, such that the total derivative of our original function w.r.t $x$ where $x = x_0$ is approximated
by the partial derivative of our equivalent statement, $f(x) = u(a, b, s) \mid_{arguments}$,  w.r.t $s$ 
where $s = x_0$. \\

>>>>

\documentclass[preprint,12pt]{elsarticle}

@@

\note[Michael]{re: I.5, next section is taken from the thesis.}
\section{Working examples of CS, AD and FD for Newton's method}
\label{sec:WorkingExamples}
In order to illustrate the complex-step technique in a simpler form than it
appears in the \emph{Peridigm} code written for this study, an example
header-only library and two example problem definition programs that use it
have been created. All the materials described here are available in the
associated project repository.

@@

\begin{figure}[tbp] \centering \includegraphics{./figs/truss.png}
\caption{Beam and spring simulated in example program.} \label{fig:TrussSchematic}
\end{figure}

@@

Table \ref{tab:ConvergenceStudy} indicates that when using AD or the 'actual'
analytically derived Jacobian, a minimum number of iterations is achieved
corresponding to a maximum convergence rate. CS and FD are similarly affected
by a grossly large step-size, while subtractive cancellation invalidates the
results for FD for very small step-sizes. Inaccuracy for large step-sizes is is
predicted by the FD and CS formulas, which require a small
step-size to allow the neglect of higher order 12 terms. The results of this
test show that step-size selection is not unimportant when using FD, but when
properly configured any of the methods may return the same maximum convergence
rate for this problem.  Overall conclusions regarding the methods that the accuracy of the
Jacobian should not be confused with the accuracy of the equilibrium
solution, since each of the methods discussed here may be equally capable given
enough iterations. Numerical measurements of accuracy besides number of iterations taken
are not possible here because there is not basis for comparison, as the methods are solving
the problem separately and producing difference Newton iterates. Additionally, while the inaccuracy of the Jacobian may
strongly affect convergence rates, this relationship was not precisely quantified here.

@@

\note[Michael]{Re: I.5, everything after this line is changed}
\section{Convergence and displacement accuracy studies} 
\label{sec:PeridigmConvergenceStudy}
%
While the previous sections profiled the per-iteration performance of each of
the methods, it was necessary to conduct convergence rate studies in order to
get a complete picture. It is necessary for an analyist to know if the
accuracy differences measured in the per-iteration tests would amount to
differences in convergence rates among the methods if they were allowed to solve problems
normally. The question is posed, given that step-size $h$ is heuristically
selected by \emph{Perdigm} for the finite difference methods and presumable
makes a good choice, do the single-iteration Jacobian accuracy results predict
the ranking of the methods in terms of convergence rate for a given study? For
brevity, convergence rate was measured for each of the methods in solving just
the largest problem of the single-core test series. As was seen in the per-iteration results,
Jacobian accuracy did not vary greatly from test to test, such that any pair of tests
from the two series could have been considered reasonably representative. As in
the validation problem shown in ~\ref{subsec:Validation}, would each of the methods return an
equal convergence rate so long as $h$ were optimally selected?

@@

\subsection{Remarks}
Based on the experimental results, it appears that for users of \emph{Peridigm}
there is a case to be made for using AD for reasons of speed. However, certain
parallel tests done in this study showed a small performance gain over AD with
the FD method.  The results showed that CS produced more accurate
tangent-stiffness matrices than CD and FD under the parameters of the tests,
however it was determined that this difference in accuracy was not sufficient
to alter convergence rate for a test problem. In order for Jacobian accuracy to
have an impact on convergence rate, and therefor serve as a predictor of
convergence rate in per-iteration results, the amount of error detected between
a supposed perfectly accurate Jacobian and the test Jacobian needs to be
greater than allowed by the conditions seen in \cite{burkeLectures}, 
reproduced in Subsection~\ref{subsec:FurtherStudies}.

@@

However, at least for the models producing the results shown here, each of the
methods returned quadratic convergence, indicating each was sufficiently
accurate. 'Marching orders' for the analyst would be to select the swiftest
method, while it would seem accuracy above what is adequate does not enhance
the convergence rate or iteration speed until either classical Newton's convergence
conditions or the approximate case conditions are violated.

The CS method was shown to be highly accurate in calculating tangent-stiffness
matrices by the standards of this study, and it is slower than other methods
for a naive implementation. In the context of future use of CS in
\emph{Peridigm}, CS holds the advantage of relying on byte-copyable data types,
while AD requires a complicated serialization and deserialization in order to
function as byte-copyable. This is fine for copying in between MPI processes,
but deserialization on an accelerator or GPU is very difficult, as this would
require compiling a version of the AD library being used that were compatible
with the accelerator or GPU being used. \note[Michael]{re:
I.3}\add[Michael]{For the reader's convenience, a collection of contiguously
stored data elements are said to be 'byte-copiable' when the the interval
in bytes of memory spanned by a data element in the collection is constant
throughout the collection. Suppose 'contiguously stored' means, stored in
memory with a minimum of unrelated infomation stored in between data
elements in the collection. Complicated objects, such as AD datatype
elements, do not have the same size in memory as other objects even of that
same class, for the reason that they themselves contain an uncertain
ammount of state data according to their design and work done in the
program. This means that a collection of such objects, even if they are
contiguously stored in memory, will not be alligned according to a constant
interval, such that elements within a collection cannot be located by
coordinates consisting of the location of the beginning of the first
element in the collection and an interval multiplied by a natural number.
This method of locating is integral to the serialized communication
implemented by MPI the libraries as well as CUDA, OpenCL, and others,
which allow only the transmission of collections of byte-copiable datatypes
(that is 'primitive datatypes'). A common method of bypassing this
limitation for communication purposes is to write the complicated data to a
'character buffer', and transmit the byte-copiable character buffer,
however this is useless unless the recipient device of the communication
can recover the original data and process it, which is impossible if the
recipient device only can work with a subset of the programming language
that the sending device must use, such as in a communication between a CPU
running C++ and a GPU running C (such as in CUDA programs), where the gap in
sophistication contains the features of the "complicated" datatype.} What this means is
that if \emph{Peridigm} or any simulation software were to be reformulated to
take advantage of accelerators or GPUs, the flexibilty of complex-step owed to
its simple implementation may make it more viable for that application given
its high accuracy. Additional performance gains may be achieved through the use
of the templated \emph{TPetra} package in \emph{Trilinos} which would allow the
use of complex data types natively, and alleviate the expense of the
dynamic-casts that were utilized in this study.  As a final comment, it should
be noted that from a code development perspective the CS method may be the
easiest of the methods to implement; It only requires one functional
evaluation, there is no heuristic choice of probe distance, and there is no
dependence on an external AD library.  From this perspective, the method may
still be an attractive method for TS matrix evaluation. \add[Michael]{However, at least for 
the models producing the results shown here, each of the
methods returned quadratic convergence, indicating each was sufficiently
accurate. 'Marching orders' for the analyst would be to select the swiftest
method, while it would seem accuracy above what is adequate does not enhance
the convergence rate or iteration speed until either classical Newton's convergence
conditions or the approximate case conditions are violated.}

@@

\item 
\label{differentiation}
%AD is equivalent to the chain rule, when the computer has definitions for functions and partials
%that can be adapted to match the composition function.
It follows from \ref{composition} that we can evaluate $\frac{d}{d x}f(x) \mid_{x_0}$ with the chain rule: \\ \\
$\frac{d}{d x}f(x) = \frac{d}{d x} \cdot g(h(k(x))) \mid_{x_0}$ \\
$\frac{d}{d x}f(x) =  \frac{d{g}}{d{h}} \cdot \frac{d{h}}{d{k}}
\cdot \frac{d{k}}{d{x}}\mid_{x_0}$ \\ \\
From the rest of \ref{given} it follows that we can approximate  $\frac{d}{d x}f(x) \mid_{x_0}$ by
specializing the computer's general forms of $u, v, w$ 
according to ~\ref{abcvalues}, with parameters $a, b$ chosen for each function and held as constant, and
evaluating, such that the total derivative of our original function w.r.t $x$ where $x = x_0$ is approximated
by the partial derivative of our equivalent statement, $f(x) = u(a, b, s) \mid_{arguments}$,  w.r.t $s$ 
where $s = x_0$. \\
